---
title: "Project 2 — Data Tidying: Electricity Production by Country"
author: "Aniss Sahraoui"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false
library(tidyverse)
library(knitr)

# Shared styling, used across all three Project 2 reports
accent   <- "#2a78d6"
accent_2 <- "#eb6834"
theme_p2 <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")
```

# Data Source

This dataset comes from the Discussion 5A post **"Total Energy Production by Country"** by Ellie
Wilser. The post uses Wikipedia's list of countries by electricity production and explains the
problem with it:

> "The data is arranged with each country as a row and columns representing the different types of
> electricity generation methods. This data is untidy because of the multi-level headers, such as
> coal and gas being grouped together under fossil fuels."

The post then asks two questions:

> "1. What country is the most sustainable? (Highest amount of renewables? Lowest amount of
> fossils? Highest percentage of renewables?)
> 2. Can the countries be grouped or clustered based on their primary energy type?"

**Source:** Wikipedia, *List of countries by electricity production*
(<https://en.wikipedia.org/wiki/List_of_countries_by_electricity_production>), which in turn draws
on Ember's electricity data.

**How it was obtained:** I scraped the first table on that page and wrote it to
[`data/electricity_production_wide.csv`](data/electricity_production_wide.csv), preserving the wide
layout. The page uses a two-row header, which a CSV cannot hold directly, so I joined the two levels
with a pipe: the column under *Fossil fuels → Coal* became `Fossil fuels | Coal`. No information is
lost, the file stays untidy in exactly the way the post describes, and the tidying step has to split
those names apart.

# Data Structure Before Tidying

```{r}
#| label: import
#| code-fold: false
raw_power <- read_csv(
  "https://raw.githubusercontent.com/AnissSahraoui/DATA607/main/Week6/data/electricity_production_wide.csv",
  show_col_types = FALSE
)

dim(raw_power)
```

```{r}
#| label: structure
names(raw_power)
```

```{r}
#| label: sample
raw_power |> slice_head(n = 5) |> select(1:7)
```

The file has **`r nrow(raw_power)` rows and `r ncol(raw_power)` columns**. It is untidy in three
distinct ways:

1. **Two variables are buried in the column names.** `Fossil fuels | Coal` encodes both an energy
   group and a specific source.
2. **Generation source is spread across columns** rather than being a value in one column.
3. **Numbers are stored as text.** Values like `10,579.70` carry thousands separators, so the
   columns import as character.

There is also a row for `World`, which is the sum of the country rows rather than a country.

```{r}
#| label: raw-types
tibble(
  column = names(raw_power),
  type   = map_chr(raw_power, ~ class(.x)[1])
) |>
  slice_head(n = 6) |>
  kable()
```

# Transformation Steps

```{r}
#| label: tidy
#| code-fold: false
power <- raw_power |>
  # 1. Reshape wide to long. names_sep splits each "Group | Source" heading into
  #    two columns, which recovers the page's two-level header as real variables.
  pivot_longer(
    cols      = contains("|"),
    names_to  = c("energy_group", "source"),
    names_sep = " \\| ",
    values_to = "generation_twh"
  ) |>
  # 2. Rename the remaining columns to snake_case for consistency.
  rename(
    country       = Location,
    country_total = `Total (TWh)`,
    year          = Year
  ) |>
  # 3. Convert text to numbers by stripping the thousands separators.
  mutate(
    across(c(generation_twh, country_total),
           ~ as.numeric(str_remove_all(as.character(.x), ","))),
    year = as.integer(year),
    # 4. Normalize an ambiguous label: "Other" appears under BOTH groups, meaning
    #    oil-fired plants in one case and geothermal etc. in the other. Left alone,
    #    grouping by source would silently merge two different things.
    source = case_when(
      energy_group == "Fossil fuels" & source == "Other" ~ "Other fossil",
      energy_group == "Renewables"   & source == "Other" ~ "Other renewable",
      TRUE                                               ~ source
    )
  )

glimpse(power)
```

**Missing values.** Blank cells are common in this table:

```{r}
#| label: missing
#| code-fold: false
power |>
  summarise(
    rows            = n(),
    missing_values  = sum(is.na(generation_twh)),
    pct_missing     = round(100 * mean(is.na(generation_twh)), 1)
  )
```

On Wikipedia a blank cell means the country generates **no measurable electricity from that
source**, not that the value is unknown: a country with no nuclear plants simply has an empty
nuclear cell. The country totals confirm this reading, since they equal the sum of the non-blank
entries. I therefore **replace blanks with 0 rather than dropping them**, because dropping the rows
would make countries look as though they had fewer sources rather than zero output from some.

```{r}
#| label: fill-zero
#| code-fold: false
power <- power |> mutate(generation_twh = replace_na(generation_twh, 0))

# Verify the decision: do the per-source values now add up to the published total?
power |>
  group_by(country, country_total) |>
  summarise(sum_of_sources = sum(generation_twh), .groups = "drop") |>
  mutate(difference = abs(sum_of_sources - country_total)) |>
  summarise(
    countries       = n(),
    max_difference  = round(max(difference, na.rm = TRUE), 2),
    within_1_twh    = sum(difference < 1, na.rm = TRUE)
  ) |>
  kable()
```

The sources add to the published total for every row, which confirms that blank means zero.

**Separating countries from the world total.** The `World` row is an aggregate, so including it in
a country ranking would be double counting. Two countries report zero generation entirely, so they
carry no mix to analyse and are excluded from percentage questions.

```{r}
#| label: split-world
#| code-fold: false
world <- power |> filter(country == "World")

countries <- power |>
  filter(country != "World") |>
  group_by(country) |>
  mutate(total_generation = sum(generation_twh)) |>
  ungroup() |>
  filter(total_generation > 0)

tibble(
  raw_rows            = nrow(raw_power),
  world_rows_removed  = 1,
  zero_output_removed = n_distinct(power$country) - 1 - n_distinct(countries$country),
  countries_analysed  = n_distinct(countries$country)
) |>
  kable()
```

# Analytical Methods

The post asks about sustainability and about grouping countries by primary energy type. I answer
them as follows:

- **Sustainability** has three readings in the post, and they give different winners, so I report
  all three: the largest *absolute* renewable output, the highest *share* of renewables, and the
  share among countries large enough for the comparison to be meaningful (over 100 TWh).
- **Grouping** uses each country's **dominant source**: the single source producing the most
  electricity there. This is a deterministic classification rather than a clustering algorithm, so
  it needs no random seed and the group labels mean something concrete.

## The world mix, for reference

```{r}
#| label: world-mix
world |>
  mutate(share = 100 * generation_twh / country_total) |>
  arrange(desc(share)) |>
  transmute(`Energy group` = energy_group, Source = source,
            `TWh` = round(generation_twh),
            `Share of world (%)` = round(share, 1)) |>
  kable()
```

## Sustainability, three ways

```{r}
#| label: sustainability
mix <- countries |>
  group_by(country, total_generation) |>
  summarise(
    renewable_twh = sum(generation_twh[energy_group == "Renewables"]),
    fossil_twh    = sum(generation_twh[energy_group == "Fossil fuels"]),
    nuclear_twh   = sum(generation_twh[energy_group == "Nuclear"]),
    .groups = "drop"
  ) |>
  mutate(
    pct_renewable = 100 * renewable_twh / total_generation,
    pct_fossil    = 100 * fossil_twh / total_generation
  )

mix |>
  slice_max(renewable_twh, n = 5) |>
  transmute(Country = country,
            `Renewable TWh` = round(renewable_twh),
            `Total TWh` = round(total_generation),
            `Renewable share (%)` = round(pct_renewable, 1)) |>
  kable(caption = "Largest renewable output in absolute terms")
```

```{r}
#| label: pct-renewable
mix |>
  filter(pct_renewable >= 99.5) |>
  arrange(desc(total_generation)) |>
  transmute(Country = country,
            `Total TWh` = round(total_generation, 1),
            `Renewable share (%)` = round(pct_renewable, 1)) |>
  kable(caption = "Countries generating essentially all electricity from renewables")
```

```{r}
#| label: fig-large
#| fig-cap: "Renewable share of generation for countries producing more than 100 TWh. Only large producers are shown, because a country generating a few TWh can reach 100% with a single dam."
#| fig-height: 4.2
large <- mix |> filter(total_generation > 100)

large |>
  slice_max(pct_renewable, n = 12) |>
  mutate(country = fct_reorder(country, pct_renewable)) |>
  ggplot(aes(x = pct_renewable, y = country)) +
  geom_col(fill = accent, width = 0.65) +
  geom_text(aes(label = sprintf("%.0f%%", pct_renewable)), hjust = 1.2,
            color = "white", fontface = "bold", size = 3.3) +
  labs(title = "Norway and Brazil lead the large producers on renewable share",
       subtitle = "Countries generating more than 100 TWh per year",
       x = "Renewable share of generation (%)", y = NULL) +
  theme_p2 +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: most-fossil
large |>
  slice_max(pct_fossil, n = 5) |>
  transmute(Country = country,
            `Total TWh` = round(total_generation),
            `Fossil share (%)` = round(pct_fossil, 1)) |>
  kable(caption = "Most fossil-dependent large producers")
```

## Grouping countries by dominant source

```{r}
#| label: dominant
dominant <- countries |>
  group_by(country) |>
  slice_max(generation_twh, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(country, energy_group, source, total_generation)

dominant |>
  count(source, name = "countries") |>
  arrange(desc(countries)) |>
  kable(caption = "Number of countries by their single largest generation source")
```

```{r}
#| label: fig-dominant
#| fig-cap: "Countries grouped by their dominant generation source, coloured by energy group."
#| fig-height: 3.6
dominant |>
  count(energy_group, source, name = "countries") |>
  mutate(source = fct_reorder(source, countries)) |>
  ggplot(aes(x = countries, y = source, fill = energy_group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = countries), hjust = -0.3, size = 3.3) +
  scale_fill_manual(values = c("Fossil fuels" = accent_2, "Renewables" = "#1baf7a",
                               "Nuclear" = accent), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Oil and hydro dominate the most countries, for opposite reasons",
       x = "Number of countries", y = NULL) +
  theme_p2 +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

```{r}
#| label: group-totals
dominant |>
  group_by(energy_group) |>
  summarise(countries = n(),
            combined_twh = round(sum(total_generation)),
            .groups = "drop") |>
  arrange(desc(countries)) |>
  kable(col.names = c("Dominant energy group", "Countries", "Their combined output (TWh)"))
```

# Conclusions

- **"Most sustainable" depends entirely on how it is measured, which is the real finding here.**
  By absolute renewable output the answer is **China**, at
  `r round(max(mix$renewable_twh))` TWh, but that is only
  `r round(mix$pct_renewable[which.max(mix$renewable_twh)], 1)`% of its own generation, because it
  also burns more coal than any other country. By share, nine countries are effectively 100%
  renewable, but most are small: the Central African Republic reaches 100% on 0.1 TWh, which is
  roughly a rounding error in China's total.
- **Among large producers, Norway is the clearest answer**, at
  `r round(large$pct_renewable[large$country == "Norway"], 0)`% renewable on 161 TWh, almost
  entirely hydro, followed by Brazil at
  `r round(large$pct_renewable[large$country == "Brazil"], 0)`% on 751 TWh. Brazil is arguably the
  more impressive case, since it sustains that share at nearly five times Norway's scale.
- **The most fossil-dependent large producers are oil and gas exporters**: Iraq (98.4%), Bangladesh
  (97.9%) and Saudi Arabia (97.8%) all generate almost nothing from renewables.
- **Countries do group cleanly by dominant source, into
  `r n_distinct(dominant$source)` groups.** The two largest are telling opposites: 64 countries run
  mostly on *other fossil fuels*, which here means oil, and these are overwhelmingly small island
  and developing states that burn imported diesel; 56 run mostly on *hydro*, usually because
  geography handed them a river. Only 7 countries are led by wind or solar, which shows how recent
  those technologies still are at national scale.
- **Fossil fuels dominate in `r sum(dominant$energy_group == "Fossil fuels")` of the
  `r nrow(dominant)` countries analysed**, and the world mix is still
  `r round(100 * sum(world$generation_twh[world$energy_group == "Fossil fuels"]) / world$country_total[1], 1)`%
  fossil-fired. The countries with the greenest mixes are mostly those with the luck of a large
  river, not those that have transitioned away from coal.

The tidying step is what made the comparison possible: the renewable share of each country requires
grouping by `energy_group`, a variable that existed only as a header spanning three columns in the
source table.
