---
title: "Project 2 — Data Tidying: Life Expectancy by Country"
author: "Aniss Sahraoui"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false
library(tidyverse)
library(knitr)

# Shared styling, used across all three Project 2 reports
accent    <- "#2a78d6"
accent_2  <- "#eb6834"
theme_p2  <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")
```

# Data Source

This dataset comes from the Discussion 5A post **"Life Expectancy by Country — World Bank"** by
Mohd Afwan Shaikh. The post describes the World Bank's *Life expectancy at birth, total (years)*
indicator, in which each country is a row and each year is a separate column, and asks:

> "I would like to compare life expectancy across different countries and examine how it has
> changed over time. I could also investigate which countries experienced the greatest improvements
> and how life expectancy changed before and after the COVID-19 pandemic."

**Source:** World Bank, indicator `SP.DYN.LE00.IN`
(<https://data.worldbank.org/indicator/SP.DYN.LE00.IN>).

**How it was obtained:** I pulled the indicator from the World Bank API for 2000–2023, kept only
sovereign countries and territories (dropping the World Bank's aggregate rows such as "World" and
"Arab World", which are sums of other rows and would be double counting), and saved the result in
the World Bank's own wide layout as
[`data/life_expectancy_wide.csv`](data/life_expectancy_wide.csv). That raw file is committed to
GitHub and is read directly from there below, so this document runs in a clean session on any
machine.

# Data Structure Before Tidying

```{r}
#| label: import
#| code-fold: false
raw_life <- read_csv(
  "https://raw.githubusercontent.com/AnissSahraoui/DATA607/main/Week6/data/life_expectancy_wide.csv",
  show_col_types = FALSE
)

dim(raw_life)
```

```{r}
#| label: structure
names(raw_life)
```

```{r}
#| label: sample
raw_life |> select(1:8) |> slice_head(n = 5)
```

The file has **`r nrow(raw_life)` rows and `r ncol(raw_life)` columns**: two identifier columns and
24 year columns.

It is untidy for one clear reason: **`year` is a variable, but it is stored in the column headings
rather than in a column.** Each row also holds 24 separate observations instead of one. In tidy
form, every row should be one country in one year, with a single `life_expectancy` value.

# Transformation Steps

```{r}
#| label: tidy
#| code-fold: false
life <- raw_life |>
  # 1. Reshape wide to long: the 24 year columns become two columns, year and value.
  pivot_longer(
    cols      = -c(`Country Name`, `Country Code`),
    names_to  = "year",
    values_to = "life_expectancy"
  ) |>
  # 2. Rename to a consistent snake_case convention. The raw names use spaces and
  #    title case, which force back-ticks everywhere in downstream code.
  rename(
    country      = `Country Name`,
    country_code = `Country Code`
  ) |>
  # 3. Normalize types: year arrives as text because it came from column names.
  mutate(year = as.integer(year)) |>
  arrange(country, year)

glimpse(life)
```

**Missing values.** The brief asks for a documented decision, so the first step is to measure the
problem rather than assume it:

```{r}
#| label: missing
#| code-fold: false
life |>
  summarise(
    rows              = n(),
    missing_values    = sum(is.na(life_expectancy)),
    countries         = n_distinct(country),
    years             = n_distinct(year),
    complete_series   = sum(!is.na(life_expectancy)) == n()
  )
```

There are **no missing values**: all `r nrow(life)` country-year combinations are present. That is
unusual for World Bank data and worth stating explicitly, because it means no imputation or
row-dropping decision is needed here. Had values been missing, I would have kept them as `NA`
rather than filling them, since an interpolated life expectancy is not an observation.

Two consistency checks on the values themselves:

```{r}
#| label: sanity
#| code-fold: false
life |>
  summarise(
    min_value      = min(life_expectancy),
    max_value      = max(life_expectancy),
    implausible    = sum(life_expectancy < 20 | life_expectancy > 100),
    duplicate_rows = sum(duplicated(across(c(country, year))))
  )
```

Every value falls in a plausible range and no country-year appears twice, so the reshaping neither
dropped nor duplicated observations.

# Analytical Methods

The post asks four questions. Each is answered from the tidy table only:

1. **How life expectancy changed over time** — the unweighted mean across countries, by year.
2. **Which countries improved most** — the change from 2000 to 2023, ranked.
3. **Before and after COVID-19** — the change from 2019 to 2021, and how many countries had
   returned to their 2019 level by 2023.
4. **How countries compare** — the spread of values in 2000 against 2023.

Averages here are **unweighted across countries**: each country counts once, regardless of
population. This answers "how are countries doing" rather than "how is the average person doing",
which is the question the post asks. A population-weighted mean would be dominated by China and
India.

## Change over time

```{r}
#| label: over-time
global <- life |>
  group_by(year) |>
  summarise(mean_life_expectancy = mean(life_expectancy), .groups = "drop")

global |>
  filter(year %in% c(2000, 2010, 2019, 2020, 2021, 2022, 2023)) |>
  mutate(mean_life_expectancy = round(mean_life_expectancy, 1)) |>
  kable(col.names = c("Year", "Mean life expectancy (years)"))
```

```{r}
#| label: fig-global
#| fig-cap: "Mean life expectancy across 217 countries, 2000–2023. The shaded band marks the COVID-19 period."
#| fig-height: 3.4
ggplot(global, aes(x = year, y = mean_life_expectancy)) +
  annotate("rect", xmin = 2020, xmax = 2021, ymin = -Inf, ymax = Inf,
           fill = accent_2, alpha = 0.12) +
  geom_line(linewidth = 1, color = accent) +
  geom_point(size = 1.8, color = accent) +
  labs(title = "Life expectancy rose for two decades, then fell sharply in 2020–21",
       x = "Year", y = "Mean life expectancy (years)") +
  theme_p2
```

The mean rose from **`r round(global$mean_life_expectancy[global$year == 2000], 1)` years in 2000**
to **`r round(global$mean_life_expectancy[global$year == 2019], 1)` in 2019**, fell to
**`r round(global$mean_life_expectancy[global$year == 2021], 1)` in 2021**, then recovered to
**`r round(global$mean_life_expectancy[global$year == 2023], 1)` by 2023**. The pandemic erased
roughly a decade of progress in two years, and the recovery took two more.

## Greatest improvements

```{r}
#| label: gains
change <- life |>
  filter(year %in% c(2000, 2019, 2021, 2023)) |>
  select(country, year, life_expectancy) |>
  pivot_wider(names_from = year, values_from = life_expectancy, names_prefix = "y") |>
  mutate(
    gain_2000_2023 = y2023 - y2000,
    covid_change   = y2021 - y2019
  )

change |>
  slice_max(gain_2000_2023, n = 5) |>
  transmute(Country = country,
            `2000` = round(y2000, 1),
            `2023` = round(y2023, 1),
            `Gain (years)` = round(gain_2000_2023, 1)) |>
  kable()
```

```{r}
#| label: fig-gains
#| fig-cap: "The ten largest gains in life expectancy, 2000 to 2023."
#| fig-height: 3.6
change |>
  slice_max(gain_2000_2023, n = 10) |>
  mutate(country = fct_reorder(country, gain_2000_2023)) |>
  ggplot(aes(x = gain_2000_2023, y = country)) +
  geom_col(fill = accent, width = 0.65) +
  geom_text(aes(label = sprintf("+%.1f", gain_2000_2023)), hjust = 1.15,
            color = "white", fontface = "bold", size = 3.4) +
  labs(title = "The largest gains are concentrated in sub-Saharan Africa",
       x = "Years gained, 2000–2023", y = NULL) +
  theme_p2 +
  theme(panel.grid.major.y = element_blank())
```

## The COVID-19 period

```{r}
#| label: covid
change |>
  slice_min(covid_change, n = 5) |>
  transmute(Country = country,
            `2019` = round(y2019, 1),
            `2021` = round(y2021, 1),
            `Change (years)` = round(covid_change, 1)) |>
  kable()
```

```{r}
#| label: recovery
recovery <- change |>
  summarise(
    countries         = n(),
    fell_2019_to_2021 = sum(covid_change < 0),
    recovered_by_2023 = sum(y2023 >= y2019)
  )

recovery |>
  kable(col.names = c("Countries", "Fell 2019 to 2021", "Back to 2019 level by 2023"))
```

## How countries compare

```{r}
#| label: spread
life |>
  filter(year %in% c(2000, 2023)) |>
  group_by(year) |>
  summarise(
    lowest  = round(min(life_expectancy), 1),
    highest = round(max(life_expectancy), 1),
    gap     = round(max(life_expectancy) - min(life_expectancy), 1),
    sd      = round(sd(life_expectancy), 2),
    .groups = "drop"
  ) |>
  kable(col.names = c("Year", "Lowest", "Highest", "Gap (years)", "Std. dev."))
```

```{r}
#| label: fig-spread
#| fig-cap: "Distribution of life expectancy across countries in 2000 and 2023."
#| fig-height: 3.2
life |>
  filter(year %in% c(2000, 2023)) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = life_expectancy, fill = year)) +
  geom_density(alpha = 0.55, color = NA) +
  scale_fill_manual(values = c("2000" = accent_2, "2023" = accent), name = NULL) +
  labs(title = "The distribution shifted right and narrowed",
       x = "Life expectancy (years)", y = "Density") +
  theme_p2 +
  theme(legend.position = "top")
```

# Conclusions

- **Life expectancy rose, fell, and recovered.** The country average climbed from
  `r round(global$mean_life_expectancy[global$year == 2000], 1)` years in 2000 to
  `r round(global$mean_life_expectancy[global$year == 2019], 1)` in 2019, dropped to
  `r round(global$mean_life_expectancy[global$year == 2021], 1)` in 2021, and reached
  `r round(global$mean_life_expectancy[global$year == 2023], 1)` in 2023, the highest in the series.
- **The biggest improvements were in sub-Saharan Africa.** Malawi gained
  `r sprintf("%.1f", max(change$gain_2000_2023))` years (46.2 to 67.4), with Rwanda, Zambia, Uganda
  and Angola each gaining more than 18. These are countries recovering from the HIV/AIDS epidemic
  and very high child mortality, so they had the most ground to make up.
- **COVID-19 hit Latin America hardest.** Bolivia lost
  `r sprintf("%.1f", abs(min(change$covid_change)))` years between 2019 and 2021, followed by
  Paraguay (−5.6), Mexico (−4.8), Guyana (−4.7) and Peru (−4.7).
  `r recovery$fell_2019_to_2021` of `r recovery$countries` countries declined over that period.
- **Most, but not all, have recovered.** `r recovery$recovered_by_2023` of `r recovery$countries`
  countries were back at or above their 2019 level by 2023, leaving
  `r recovery$countries - recovery$recovered_by_2023` still below it.
- **Countries are converging.** The gap between the highest and lowest country narrowed from 37.3
  years in 2000 to 31.9 in 2023, and the standard deviation fell from 9.64 to 7.12. The lowest
  country in 2023, Nigeria at 54.5 years, is nearly ten years above the lowest in 2000.

The practical lesson for the tidying step is that the single `pivot_longer()` call is what makes
all of this possible: every question above is a `group_by()` on a column that did not exist in the
source file.
