---
title: "Project 2 — Data Tidying: State Unemployment"
author: "Aniss Sahraoui"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false
library(tidyverse)
library(knitr)

# Shared styling, used across all three Project 2 reports
accent   <- "#2a78d6"
accent_2 <- "#eb6834"
theme_p2 <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")
```

# Data Source

This dataset comes from the Discussion 5A post **"U.S. Bureau of Labor Statistics Table"** by Zaina
Hassan, who describes the problem with it:

> "I consider this dataset untidy because the dates and measurements are spread across multiple
> columns and levels of column headings rather than each variable having its own individual column.
> ... This wide format makes the table fairly easy for a person to read on the webpage, but it would
> need to be reorganized before it could be easily analyzed in R."

and states the analysis:

> "One analysis I could perform after tidying the dataset would be to compare unemployment rates
> between states and examine how they change over time. I could also identify which states
> experienced the largest increases or decreases in unemployment during the time period represented
> in the table."

**Source:** U.S. Bureau of Labor Statistics, *State and Regional Unemployment*, Table 1:
"Employment status of the civilian noninstitutional population 16 years of age and over by region,
division, and state, 2023-24 annual averages"
(<https://www.bls.gov/news.release/srgune.t01.htm>).

**How it was obtained:** The BLS blocks automated downloads, so I transcribed the published table
into [`data/state_unemployment_wide.csv`](data/state_unemployment_wide.csv), keeping its original
structure and formatting. The page uses a two-level header, so each heading pair is joined with a
pipe: the column under *Civilian labor force → 2023* became `Civilian labor force | 2023`. I also
kept the thousands separators and the footnote marker exactly as printed, so the cleaning work is
real rather than done in advance.

I checked the transcription by verifying that *employed + unemployed = civilian labor force* in
every row, which it does to within one thousand (the table is rounded to thousands).

# Data Structure Before Tidying

```{r}
#| label: import
#| code-fold: false
raw_jobs <- read_csv(
  "https://raw.githubusercontent.com/AnissSahraoui/DATA607/main/Week6/data/state_unemployment_wide.csv",
  show_col_types = FALSE
)

dim(raw_jobs)
```

```{r}
#| label: structure
names(raw_jobs)
```

```{r}
#| label: sample
raw_jobs |> slice_head(n = 5) |> select(1:6)
```

The file has **`r nrow(raw_jobs)` rows and `r ncol(raw_jobs)` columns**, and is untidy in four
ways:

1. **Two variables sit in each column name**: a measure (population, labor force, employed,
   unemployed, rate) and a year.
2. **Each row holds ten observations**, one per measure-year combination.
3. **Numbers are text**: values such as `266,942` carry thousands separators.
4. **Rows mix three levels of geography.** The United States total, four regions, nine divisions,
   the 50 states plus DC, and Puerto Rico all sit in one column. Summing that column would count
   every person several times over.

```{r}
#| label: raw-types
tibble(column = names(raw_jobs), type = map_chr(raw_jobs, ~ class(.x)[1])) |>
  slice_head(n = 6) |>
  kable()
```

# Transformation Steps

```{r}
#| label: tidy
#| code-fold: false
# The region and division names printed in the table's stub column. Listing them
# explicitly is the only reliable way to tell a region row from a state row.
regions <- c(
  "United States",
  "Northeast", "New England", "Middle Atlantic",
  "Midwest", "East North Central", "West North Central",
  "South", "South Atlantic", "East South Central", "West South Central",
  "West", "Mountain", "Pacific"
)

jobs <- raw_jobs |>
  # 1. Rename to snake_case. The first column's name is a sentence in the source.
  rename(
    area        = `Region, division, and state`,
    error_range = `Error range of rate, 2024`
  ) |>
  # 2. Reshape wide to long, splitting each "Measure | Year" heading into two
  #    columns. This is what turns the two-level header into real variables.
  pivot_longer(
    cols      = contains("|"),
    names_to  = c("measure", "year"),
    names_sep = " \\| ",
    values_to = "value"
  ) |>
  # 3. Normalize types and labels.
  mutate(
    value   = as.numeric(str_remove_all(value, ",")),   # "266,942" -> 266942
    year    = as.integer(year),
    measure = str_replace_all(str_to_lower(measure), " ", "_"),
    # 4. Add the geography level that was implicit in the row ordering, so the
    #    three levels can be separated instead of being summed together.
    area_type = case_when(
      area %in% regions ~ "region_or_division",
      area == "Puerto Rico" ~ "territory",
      TRUE ~ "state"
    )
  )

glimpse(jobs)
```

```{r}
#| label: units
#| code-fold: false
# Record the units, which the source gives only in a note above the table.
units <- tribble(
  ~measure,              ~unit,
  "population",          "thousands of people",
  "civilian_labor_force","thousands of people",
  "employed",            "thousands of people",
  "unemployed",          "thousands of people",
  "unemployment_rate",   "percent of labor force"
)

units |> kable(col.names = c("Measure", "Unit"))
```

**Missing and inconsistent values.** Two issues appear, and they need different treatment:

```{r}
#| label: missing
#| code-fold: false
jobs |>
  summarise(
    rows                   = n(),
    missing_numeric_values = sum(is.na(value))
  ) |>
  kable()

# The error range column is text, and one row carries a footnote marker instead of a range.
jobs |>
  distinct(area, error_range) |>
  filter(str_detect(error_range, "\\(")) |>
  kable()
```

Every numeric value is present. The one inconsistency is in `error_range`, where Puerto Rico shows
`(2)`, a footnote meaning "data not available", rather than a range like `3.9 - 4.1`. I convert that
marker to a proper `NA` rather than leaving the string in place, because `(2)` is not a value and
would otherwise be silently treated as one:

```{r}
#| label: footnote
#| code-fold: false
jobs <- jobs |>
  mutate(error_range = if_else(str_detect(error_range, "\\("), NA_character_, error_range))

sum(is.na(jobs$error_range)) / n_distinct(jobs$measure) / n_distinct(jobs$year)
```

**A check on the reshaping.** The unemployment rate should equal unemployed divided by labor force.
Recomputing it from two other columns confirms the long table holds together:

```{r}
#| label: check
#| code-fold: false
jobs |>
  filter(measure %in% c("unemployed", "civilian_labor_force", "unemployment_rate")) |>
  pivot_wider(names_from = measure, values_from = value) |>
  mutate(recomputed = 100 * unemployed / civilian_labor_force,
         difference = abs(recomputed - unemployment_rate)) |>
  summarise(
    rows           = n(),
    max_difference = round(max(difference), 2),
    median_difference = round(median(difference), 3)
  ) |>
  kable()
```

The largest gap is 0.17 percentage points and the median is 0.03, which is what rounding to
thousands of people produces. The published rates are computed from unrounded data, so exact
agreement is not expected.

# Analytical Methods

The post asks to compare states, look at change over time, and find the largest moves. Because the
table covers two annual averages (2023 and 2024), "over time" means the change between those two
years.

All analysis uses the **51 state rows (50 states plus the District of Columbia)**. Region, division
and national rows are excluded from comparisons and rankings, since they are aggregates of the same
people; Puerto Rico is excluded because its data comes from a different survey, which the source
note states explicitly. The national figure is used only as a reference line.

```{r}
#| label: rates
#| code-fold: false
state_rates <- jobs |>
  filter(area_type == "state", measure == "unemployment_rate") |>
  select(area, year, rate = value) |>
  pivot_wider(names_from = year, values_from = rate, names_prefix = "y") |>
  mutate(change = y2024 - y2023)

us_rate <- jobs |>
  filter(area == "United States", measure == "unemployment_rate") |>
  select(year, rate = value)

us_rate |> kable(col.names = c("Year", "U.S. unemployment rate (%)"))
```

## Comparing states

```{r}
#| label: extremes
bind_rows(
  state_rates |> slice_max(y2024, n = 5) |> mutate(group = "Highest in 2024"),
  state_rates |> slice_min(y2024, n = 5) |> mutate(group = "Lowest in 2024")
) |>
  transmute(Group = group, State = area,
            `2023 (%)` = y2023, `2024 (%)` = y2024,
            `Change (pp)` = sprintf("%+.1f", change)) |>
  kable()
```

```{r}
#| label: fig-states
#| fig-cap: "Unemployment rate by state, 2023 and 2024 annual averages. The dashed line is the 2024 national rate."
#| fig-height: 7.5
state_rates |>
  mutate(area = fct_reorder(area, y2024)) |>
  ggplot(aes(y = area)) +
  geom_vline(xintercept = us_rate$rate[us_rate$year == 2024],
             linetype = "dashed", color = "grey55") +
  geom_segment(aes(x = y2023, xend = y2024, yend = area), color = "grey70", linewidth = 0.7) +
  geom_point(aes(x = y2023, color = "2023"), size = 2) +
  geom_point(aes(x = y2024, color = "2024"), size = 2) +
  scale_color_manual(values = c("2023" = accent_2, "2024" = accent), name = NULL) +
  labs(title = "Unemployment rose in almost every state between 2023 and 2024",
       subtitle = "Dashed line: 2024 national rate of 4.0%",
       x = "Unemployment rate (%)", y = NULL) +
  theme_p2 +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
```

## Largest increases and decreases

```{r}
#| label: movers
bind_rows(
  state_rates |> slice_max(change, n = 5) |> mutate(group = "Largest increases"),
  state_rates |> arrange(change) |> slice_head(n = 5) |> mutate(group = "Smallest changes")
) |>
  transmute(Group = group, State = area,
            `2023 (%)` = y2023, `2024 (%)` = y2024,
            `Change (pp)` = sprintf("%+.1f", change)) |>
  kable()
```

```{r}
#| label: direction
state_rates |>
  summarise(
    states          = n(),
    mean_2023       = round(mean(y2023), 2),
    mean_2024       = round(mean(y2024), 2),
    rose            = sum(change > 0),
    fell            = sum(change < 0),
    unchanged       = sum(change == 0)
  ) |>
  kable(col.names = c("States", "Mean 2023 (%)", "Mean 2024 (%)",
                      "Rose", "Fell", "Unchanged"))
```

```{r}
#| label: fig-change
#| fig-cap: "Change in unemployment rate from 2023 to 2024, in percentage points."
#| fig-height: 7.5
state_rates |>
  mutate(area = fct_reorder(area, change),
         direction = if_else(change > 0, "Rose", "Fell or unchanged")) |>
  ggplot(aes(x = change, y = area, fill = direction)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Rose" = accent_2, "Fell or unchanged" = accent), name = NULL) +
  labs(title = "Only three states saw unemployment fall",
       x = "Change in unemployment rate, 2023 to 2024 (percentage points)", y = NULL) +
  theme_p2 +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
```

# Conclusions

- **Unemployment rose almost everywhere.** Of the `r nrow(state_rates)` states,
  `r sum(state_rates$change > 0)` saw their rate rise between 2023 and 2024, only
  `r sum(state_rates$change < 0)` saw it fall, and `r sum(state_rates$change == 0)` were unchanged.
  The state average went from `r round(mean(state_rates$y2023), 2)`% to
  `r round(mean(state_rates$y2024), 2)`%, alongside a national move from 3.6% to 4.0%.
- **The spread between states is wide.** In 2024 the highest state rate, Nevada at
  `r max(state_rates$y2024)`%, was more than three times the lowest, South Dakota at
  `r min(state_rates$y2024)`%. California (5.3%), the District of Columbia (5.2%), Kentucky (5.1%)
  and Illinois (5.0%) round out the top five, while the Plains and northern New England states sit
  at the bottom.
- **The largest increases were not in the states with the highest rates.** Rhode Island rose the
  most, by `r sprintf("%+.1f", max(state_rates$change))` points, from 3.0% to 4.3%, followed by
  South Carolina (+1.1) and Colorado (+1.0). All three started below the national average and ended
  near or above it, so the increase reflects a genuine turn rather than an already-weak labor
  market getting worse.
- **The three declines were marginal.** Pennsylvania, Arizona and Delaware each fell by 0.1 points,
  which is within the error ranges the BLS publishes for these estimates, so they are best read as
  flat rather than as improvements.
- **A caution the table itself supplies.** The published error range for 2024 is roughly ±0.2 to
  ±0.6 points for most states. Differences smaller than that, including every one of the three
  declines, should not be treated as real changes.

The tidying step is what enabled this: the comparison of 2023 against 2024 requires `year` to be a
variable, and in the source it existed only as a second row of column headings.
